Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-20 paper/reviewer_response_PNAS.tex

\subsubsection*{Comments:} 
In their manuscript Lee et al. examine the tolerance of an H3 virus HA for single point mutations using deep mutational scanning. Using the data obtained from cell culture experiments the authors then try to understand if there is a correlation between the observed changes in vitro and actual virus evolution in vivo. Finally, they test if there is any correlation between H3 and H1 mutational tolerance and basically find that measurements made with H1 HA are not helpful to predict H3 evolution. The manuscript is written in a relatively confusing way and the experimental setup is problematic. Some of the obtained data also need re-evaluation/experimental validation.

\subsubsection*{Major points}

1) It is disturbing that there is no preference for having a methionine in the first position of the HA ORF. The authors discuss this and refer to one 2002 paper that shows additional start sites for non-human H2. This is a pretty weak explanation. If this is really the case the authors should actually isolate/rescue these viruses and test their fitness relative to wild type.

\response{We have now rescued by reverse genetics influenza GFP viruses carrying the M(-16)K mutation in the start codon, and titered by flow cytometry. 
By 53 hours post-transfection when we harvested transfection supernatant, the M(-16)K mutant reached a titer of approximately one log less than that of wildtype virus.
Although we may have expected the M(-16)K mutant to grow comparably to wildtype given that Met and Lys are both preferred at the start codon, we would like to emphasize that (1) we grew influenza GFP viruses which may not behave identically to live influenza viruses and (2) we did not capture the full viral growth kinetics of this mutant since we titered at a relatively early timepoint after transfection.
In comparison, the C52A variant, which mutates the highly conserved Cys at site 52, grows very poorly, while the synonymous C52C mutant grows to titers comparable to those of wildtype.
These results are now shown in Figure S9, and we have updated the Methods section describing how we validated these point mutants.
}

@@

Of note, position 1 for the H1 HA in the supplementary figure is not even shown.

\response{Figure~S3 shows data for the H1 HA as measured previously by Doud and Bloom (2016). 
That study used mutant libraries in which all codons \emph{except} the one for the N-terminal methionine were mutagenized.
Therefore, that study did not estimate the effects of mutations to the first codon, which is why it is not shown in Figure~S3.
We have added a sentence to the legend for this supplementary figure explaining this fact.
}

2) While other important features of the HA like several cysteines seem to be conserved, there is no pressure on other elements which are considered important for viral fitness including the NGT glycosylation site in the beginning of HA1 which is conserved in all group 2 HAs or the adjacent NAT glycosylation site which is conserved in H3, H7, H10 and H15. On the other hand, head glycosylation sites, which are highly variable in nature, seem to be conserved. Again, viruses without these glycosylation sites should be rescued and evaluated for fitness relative to wild type.

\response{This is a valid point, and we have also individually evaluated the growth of three glycosylation site mutants.
Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites, respectively.
The third mutant, S287A, removes a constrained glycosylation site.
In Figure S9, we show that the T24F and T40V mutants grow to titers comparable to or higher than that of wildtype. 
However, the titer of the S287A is approximately 2.8-fold lower than that of wildtype, suggesting that the glycan at sites 285-287 is important for viral growth in cell culture.

Furthermore, constraint in our experiments and conservation in nature are not expected to be entirely the same.
Conservation in nature results from the convolution of constraint and diversifying selection from immunity, whereas our experiments only look at constraint.
\comment{As evidence to explain this point, we can point to studies showing that the head of HA is much more conserved for swine influenza viruses where there is less pressure from immunity.}
}

\comment{Alanine is not the least preferred amino acid at site 287, while it is the least preferred at site 52.}

@@

3) It is also surprising that the transmembrane domain seems to be highly plastic, a feature that is certainly not present in nature where it is highly conserved within H3 HAs.

\response{We validated growth of a transmembrane domain mutant, C199(HA2)K, and show that this mutant attains titers approximately 1.6-fold higher than that of wildtype.
In addition, various studies have found that chimeric HA's with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB), and that the length of the transmembrane domain seems to play a large role in determining HA fusion activity (Armstrong et al 2000 JCB).
}

@@

4) Were sequences from passaged viruses (isolated on MDCKs and other cell lines) included in the comparison between in vitro mutagenesis data and natural evolution data? They need to be excluded since their presence might induce a bias towards what is tolerated/preferred in cell culture. Only sequences obtained directly from patient specimens should be used for the comparison.

\response{We thank the reviewer for this astute suggestion and now include an analysis of the relationship between mutational effect and maximum frequency for mutations from unpassaged isolates.
Due to the sparseness of unpassaged isolate sequences previous to the last decade, we analyzed unpassaged isolates from the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
Figures 5, 6, and S7 now include panels showing the results for these unpassaged isolate mutations.
The results in Figure 5 demonstrate that the correlation between mutational effect as determined by our experiments and maximum mutation frequency is stronger for unpassaged isolates than when we include all isolates ($\rho = 0.24$).
We do not see such a correlation when we use the WSN/1933 H1 HA preferences to estimate the mutational effects for unpassaged isolate mutations.
}

@@

5) It is unclear how the ``rescaling'' is performed and what it exactly does. The authors need to described and justify this better and should also provide the non-rescaled data in a ``sequence logo'' figure for comparison.

\response{Rescaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
This parameter is fit under an experimentally informed codon substitution model (ExpCM) using the phydms program (Hilton 2017 PeerJ).
A stringency parameter greater than 1 indicates that natural selection favors the same amino acids as the experiments but with greater stringency, while a value less than 1 indicates that natural selection does not favor the amino acids with as much stringency as the experiments.
Because selection in the lab often does not fully reflect the stringency of selection in nature and selection across experiments may not be entirely consistent, we re-scale the preferences by this stringency parameter.
Importantly, re-scaling does not affect any of the evolutionary analyses performed in this paper since it is a constant that multiplies all mutational effects, simply corresponding a change in units on the x-axis of Figure~5.
We now include a supplementary panel A to Figure S3 showing the un-scaled Perth/2009 H3 HA preferences.
}

@@

6) The authors state that it there are more reports of escape from stalk antibodies for H3 than H1. This is simply not true. Actually, the first reference used as evidence for easy selection of H3 anti-stalk escape mutants (Chai et al) reported three escape mutants which a) did not disrupt binding completely and b) were crippled in vitro and in vivo. The authors should remove their statement as there is not systematic data to support this.

\response{We have now removed this sentence from the paper.}

@@

\comment{We looked at this pretty carefully, and there are definitely more reports of selecting anti-stalk antibodies for H1 and H3.
The reviewer's counterpoint is to mention on of the H3 studies. 
However, we didn't say that there are no H3 reports, just that there are a lot more H1 reports.
Certainly this is not systematic proof, but our statement in the text goes to great pains to caveat that this is purely circumstantial and not systematic proof.
But since this isn't a central claim of the paper of anyway, let's just remove it in a concession to the reviewer.
The study we're helping Nicholas Wu with will eventually systematically test this point.
}

7) The biggest issue with the manuscript is the experimental set up. The authors titrate their rescued virus using a TCID50 assay on MDCK cells. Based on the titer that they measure, they passage the virus from the initial rescue once onto fresh MDCK cells using a low MOI. After infection, the virus dilution stays on the cells and is then harvested without centrifugation followed by RNA extraction and sequencing. There are many issue with this process.

\response{The reviewer has a number of questions and suggestions about the experimental design.
Over the last few years, we have performed a number of studies optimizing procedures for deep mutational scanning of influenza virus libraries (see PMIDs 24859245, 25006036, 27271655) during which we have carefully considered all of the experimental parameters that the reviewer discusses.
Below we briefly explain why each of our choices for these experimental parameters were appropriate.
We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
\begin{itemize}
\item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
\item The experimental measurements do a good job of predicting the effects of individually validated mutations, as shown in Figure S8.
\item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
\item The experimental measurements provide information that can help predict the fates of influenza virus lineages in nature (Fig.~5).
\end{itemize}
We do not claim that other experimental parameter choices might not lead to equivalent or even slightly better results, but we have carefully considered all of our choices and clearly they lead to very good results.

@@

Below we explain our reasoning for each choice about which the reviewer makes suggestions:
}

a) The TCID50 is likely underestimating the number of virus particles with HA genomic segments because not all of them (likely the minority) will be capable of inducting a full infectious cycle (but their presence might still be detected in the cell supernatants).

\response{The reviewer is certainly correct that TCID50 assays underestimate the number of HA genomic segments, as it is well known that the number of physical influenza virions exceeds the number of infectious units by 10-100 fold.
This is why we used an extremely low MOI (0.0035) so that most cells are still infected by no more than one virion.
It is impossible to ever grow a viral stock that is completely free of defective particles.
However, in both our original rescue with helper virus and in the passage, we follow the steps suggested by Carolina Lopez and Scott Hensley (PMID 27047455) to minimize defective particles.
These include low MOI (0.0035 TCID50 per cell is quite low), short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours), and replacing the media three hours post-infection to remove residual defective particles.
}

@@

b) The infectious inoculum is never washed off the infected cells meaning that any particles that are transferred from the rescue transfection are going to be detected during sequencing. This might even apply to naked RNPs released from dead cells.

\response{For biological replicates 1, 2, and 3-2, we replaced the original infectious inoculum with 25 mL fresh IGM three hours post-infection.
For replicate 3-1, we did not change the media. 
However, for this replicate, the original inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing.
Specifically, in the original inoculum placed on the cells, the titer is 2.5 TCID50 / ul, whereas at the time of supernatant collection for sequencing the titer is 32 TCID50 / ul for unconcentrated virus but 400 TCID50 / ul for the concentrated virus.
Therefore, the original inoculum constitutes just 8\% of the viral titer at the time of collection, and therefore makes a negligible contribution to the sequenced material.
We deeply apologize for not including this crucial step in the Methods, and have updated the Methods to now include completion of this step. 
We have also included in the GitHub repository a PDF of the lab notebook describing the rescue and passage of the mutant virus libraries at %\url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09%20HA%20Library%20Rescue.pdf} (relevant details of the notebook can be found in pages 20-25).
}

@@

c) This would actually not be an issue, if the dilution from the rescue culture to the new culture would be large enough. But this is not the case. The authors passaged $9\times10^5$ TCID50 (and likely much more DIs/RNA) from the rescue cultures onto the new cells in a volume of 25 ml. This is, depending on experiment, between 10 and 1 ml of supernatant of the original rescue and leads to a 1:2.5 or 1:25 dilution respectively.

\response{We passaged for each mutant virus library $9\times10^5$ TCID50 total across 15 $\times$ 15-cm dishes.
Because we used an inoculum of 2.5 TCID50 / $\mu$l across these fifteen dishes, this corresponds to dilutions that range from 1:37 to 1:294.
The quantity of dishes used to passage each mutant virus library is now mentioned in the Methods.
}

@@

d) The supernatant is never centrifuged before RNA extraction. Cell debris is not removed which is likely contributing significantly to RNA sequences detected in the supernatant. However, RNA associated with cell debris is not representative of RNA from fit, replicating viruses.

\response{We clarified the viral supernatant before RNA extraction by centrifuging the supernatant at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris.
We failed to mention this step in the Methods, and again deeply apologize for this critical oversight.
We have updated the Methods to now include mention of this step.
}

@@

\response{For the reasons described above, it is not necessary to totally repeat the experiments.
Certainly other experimental parameter choices could have been used, but this is always the case for complex experiments.
Our choices will reasonable, and the four lines of evidence listed at the beginning of this comment show that they yield reproducible and meaningful results.}

\subsubsection*{Minor points}

1) Many abbreviations are not defined in the manuscript including MDCK-SIAT1, TMPRSS2, MOI, DMEM, FBS, LB.

\response{We now explain these abbreviations in the manuscript.}

@@

2) The authors state on page 5 that mutations detected in cell culture as deleterious rarely reach high frequency in nature. Could you please give a few examples?

\comment{Not obvious how to give an example, as there are many mutations we identify as deleterious and many mutations that don't reach high frequency. 
But I suppose if we validate some specific mutation (e.g., a cysteine mutation) as deleterious, we can point out that it doesn't reach high frequency...}

3) The authors picked 31 random clones to evaluate the mutation rate per clone. These sequences should be made available. Did this set include viruses without a proper start codon (see above)? What was their growth phenotype in vitro?

\response{We have now added the list of mutations from the 31 clones to Dataset S3.
Two of the clones had a mutation that removed the ATG start codon.
}

\comment{Not sure how to address the growth phenotype question}.

@@

\subsection*{Reviewer \#2 Comments}

\subsubsection*{Comments:} 
The work presented in this manuscript is a thorough analysis of the tolerance of an H3 hemagglutinin to all possible single amino-acid mutations. H3 evolution is of particular relevance now in light of the most recent flu season where vaccine efficacy against H3N2 was estimated at only 25\%. The experimental setup presented in the manuscript is sound and there is strong correlation among the biological and technical replicates. The authors show that data from deep mutational scanning is better able to describe the evolution of H3 HA compared to conventional models. This is consistent with the results of their previous study using deep mutational scanning on the H1 HA of WSN. It is important to note that their approach is ``systematic and quantitative'' in capturing potential effects of mutations in the H3 HA on viral fitness in cell culture. The significance statement is accurate in describing the conclusions of the paper and is understandable to a general scientific audience.

\response{We thank the reviewer for an accurate and fair summary of our work.}

@@

\subsubsection*{Major points:}

The authors themselves acknowledge the most important limitation of this study: that the experimental measure of viral growth in cell culture is not reflective of true fitness in nature. Most importantly, the work presented here does not capture host-induced pressures on the antigenic regions of the viral protein. This is evidenced by the discrepancy between the high mutational tolerance they calculated for the stalk region and the actual change in stalk amino acid sequence. A simple protein BLAST comparison between the amino acid sequences of HA from H3N2 viruses Perth09 and HK68 shows a 77\% conservation in head amino acids but a 94\% conservation in stalk amino acids. Additionally, according to figure 2, there is no observable reversion of the G78D mutation they introduced despite the dominance of G78 in the population as seen in figure S1. The authors recognize this in their work and do not attempt to exaggerate the significance of these findings. Still, it would be useful to differentiate between head and stalk mutations when depicting individual amino acid mutations in figure 4 and their frequency trajectories to highlight that the predominance of mutations that occur in nature are in the head.

\response{We thank the reviewer for acknowledging our efforts to fairly present the considerable caveats associated with cell-culture measurements.

The suggestion to make a version of Fig.~4 that distinguished between the stalk and the head is a good one.
It is difficult to also overlay the structural location (stalk versus head) on the current Fig.~4, since colors are already used to indicate mutational effects.
We have therefore created a new supplemental figure (Fig.~S7) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.

@@

The reviewer is correct that G78D appears to be a site where the correspondence between selection in our experiments and frequency in nature is low.
This makes sense, as G78D is one of the two sites that we specifically identified as being a lab adaptation (see Figure~S1).
We now explicitly mention this fact in the text \comment{still need to actually add this mention}.
}

The data regarding greater mutational tolerance in the stalk compared to the head should be paired with data from deep mutational scanning utilizing stalk- and head-specific antibodies as a selection process as this group has previously done with WSN H1 (Doud et al., Nat Comm 2018, Doud et al., PLoS Path 2017) to see if tolerant sites are responsible for viral escape.

\response{We agree that this would be a very interesting experiment to perform.
However, we believe such a study, which would require rigorous examination of escape from multiple head- and stalk-specific antibodies, is beyond the scope of this manuscript.
}

@@

Additionally, the authors should directly show the discrepancies between this in vitro selection and the H3 HA's evolution in nature. This group has performed such an analysis in their previous deep mutational scanning work with the H1 HA of WSN (Doud and Bloom, Viruses 2016). Maximum likelihood phylogenetics can be used to estimate the difference between experimental amino acid preference and natural preference. One might expect that this would provide some insight into the nature of the bias introduced through cell culture.

\response{This is a good suggestion and we have now added the results of this analysis in Figure S4.
}

\comment{Also need to add mention of this figure in the manuscript text somewhere...}

@@

The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as ``modest.'' One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent. The authors should address the possibility that data from deep mutational scanning can be generalized to divergent strains if the study is done in a system that is more reflective of nature.

\response{We now have acknowledged this possibility in the Discussion in the sentence "We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.}

### 2018-06-20 paper/manuscript.tex
Update supp figures & paper text, add to reviewer responses

We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.

After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B).
We also passaged and deep sequenced library 3 in duplicate (library 3-1 and 3-2) to gauge experimental noise \textit{within} a single biological replicate.
As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.

@@

We performed a similar analysis using our new data for the Perth/2009 H3 HA.
Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
}
\end{figure}

@@

We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL (no trypsin was added since there was TMPRSS2), and then collected the viral supernatant at 72 hours post-transfection.
This viral supernatant was blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
We isolated viral RNA from these passaged viruses and sequenced the HA gene.
The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
For all subsequent experiments, we used viruses with the HA containing these two mutations to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
We used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.

The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
We created three independent libraries, one for each biological replicate.
The mutant variants were then cloned at high efficiency into the pICR2~\cite{ashenberg2017deep} vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
We obtained $>$6 million transformants for each replicate.
We scraped the plates, expanded the cultures in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).

@@

We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
Three hours post-infection, we replaced the inoculum with fresh IGM.
We collected viral supernatant for sequencing 48 hours post-infection.

@@

To determine the effect of virus passaging on the relationship between mutational effects and maximum frequencies in natural populations, we repeated the above analyses using only viruses that were sequenced directly without passaging.
Routine direct sequencing did not begin until the early 2000s \cite{McWhite2016}.
To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.

@@

@@

@@

>>>>

We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
To further increase viral titers, we used MDCK-SIAT1 (Madin-Darby canine kidney cells with enhanced expression of sialic acids) cells that we had engineered to constitutively express the TMPRSS2 (Transmembrane Protease, Serine 2) protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.

After generating the mutant virus libraries, we passaged them at low multiplicity of infection (MOI) in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B).
We also passaged and deep sequenced library 3 in duplicate (library 3-1 and 3-2) to gauge experimental noise \textit{within} a single biological replicate.
As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.

@@

We performed a similar analysis using our new data for the Perth/2009 H3 HA.
Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The Perth/2009 strain is labeled with a star, and nodes in the clade containing the Perth/2009 strain were excluded from our analyses.
The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
}
\end{figure}

@@

We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM, or Dulbecco's Modified Eagle Medium, supplemented with 10\% heat-inactivated FBS, or Fetal Bovine Serum, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL (no trypsin was added since there was TMPRSS2), and then collected the viral supernatant at 72 hours post-transfection.
This viral supernatant was blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
We isolated viral RNA from these passaged viruses and sequenced the HA gene.
The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
For all subsequent experiments, we used viruses with the HA containing these two mutations to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
We used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.

The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
We created three independent libraries, one for each biological replicate.
The mutant variants were then cloned at high efficiency into the pICR2~\cite{ashenberg2017deep} vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
We obtained $>$6 million transformants for each replicate.
We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).

@@

We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
We collected viral supernatant for sequencing 48 hours post-infection.

@@

To determine the effect of virus passaging on the relationship between mutational effects and maximum frequencies in natural populations, we repeated the above analyses using only viruses that were sequenced directly without passaging.
Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.

@@

\subsection*{Validation of individual point mutants}
To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, C199(HA2)K, T24F, T40V, and S287A, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmid and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
We changed the media in each well with 2 mL IGM eight hours post-transfection.
At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.

To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
Four hours after plating, we infected cells with dilutions of viral supernatant.
At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
\caption{\label{suppfig:diffprefs_logoplot}
{\bf The differential preferences of the Perth/2009 H3 HA compared to amino-acid frequencies from a sequence alignment of natural human H3.}
The differential preferences estimated by comparing the Perth/2009 H3 HA preferences to amino-acid frequencies from a sequence alignment of natural human H3N2 HA sequences.
For each site, the height of each letter above and below the line indicate how differentially preferred that amino acid is in our experiments compared to in nature.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
The sites are in H3 numbering.
}
\end{suppfigure}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.5\textwidth]{figs/supp_head_stalk_mutfreq/head_stalk_mutfreq.pdf}}
\caption{\label{suppfig:head_stalk_mutfreq_supp}
{\bf Frequency trajectories of head and stalk domain mutations.}
(A) This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:frequency_trajectory}, but colors amino-acid mutations by whether they occur in the head (purple) or the stalk (blue) domain.
(B) Histogram of mutation maximum frequencies by the number of nodes for head and stalk domain mutations.
It is clear that mutations in the head domain are more numerous than those in the stalk.
}
\end{suppfigure}
